---
title: "Berries Project"
author: "Anna Cook"
date: "10/19/2020"
output: ioslides_presentation
---

```{r setup, include=FALSE, warning=FALSE, message=FALSE}
knitr::opts_chunk$set(echo = FALSE)
library(knitr)
library(tidyverse)
library(magrittr)
library(kableExtra)
library(dplyr)
library(ggplot2)
```

## Introduction

- Examining how different aspects of the production of raspberries is related to their value
- Data collected from USDA database
- Dataset contains info for 3 types of berries, but we are focused only on raspberries

## Data Cleaning, Step 1

Eliminate unnecessary variables. We are left with 8 variables of interest.
```{r, echo=FALSE, warning=FALSE, message=FALSE}
ag_data <- read_csv("berries.csv", col_names = TRUE)
ag_data %>% summarize_all(n_distinct) -> aa
bb <- which(aa[1,]==1)
cn <- colnames(ag_data)[bb]
ag_data %<>% select(-all_of(bb))
aa %<>% select(-all_of(bb)) 
ag_data %<>% select(-4)
aa %<>% select(-4) 
kable(head(ag_data)) %>% kable_styling(font_size=12)
```

## Data Cleaning, Step 2

Filter the data so it displays only raspberries for each year.
```{r, echo=FALSE, warning=FALSE, message=FALSE}
rberry <- ag_data %>% filter((Commodity=="RASPBERRIES") & (Period=="YEAR"))
rberry %<>% select(-c(Period, Commodity))
kable(head(rberry)) %>% kable_styling(font_size = 12)
```

## Data Cleaning, Step 3

Split, re-organize, and re-name the Data Item and Domain Category Variables so that they are more informative. Remove redundancy by removing Domain column. 
```{r, echo=FALSE, warning=FALSE, message=FALSE}
rberry %<>% separate(`Data Item`, c("Type", "Production"), sep = "-")
rberry %<>% separate(Type, c("Berry", "Type", "Market"), sep = ",")
rberry %<>% separate(Production, c("Production", "Measure", "Average"), sep = ",")
rberry[is.na(rberry)] <- ""
rberry$Market <- ifelse(rberry$Type==" PROCESSING ", paste(rberry$Type, rberry$Market, sep = ""), rberry$Market)
rberry$Type <- ifelse(rberry$Type==" PROCESSING ", NA, rberry$Type)
rberry[is.na(rberry)] <- ""
rberry$Market <- ifelse(rberry$Type==" UTILIZED ", paste(rberry$Type, rberry$Market, sep = ""), rberry$Market)
rberry$Type <- ifelse(rberry$Type==" UTILIZED ", NA, rberry$Type)
rberry[is.na(rberry)] <- ""
rberry$Market <- ifelse(rberry$Type==" FRESH MARKET ", paste(rberry$Type, rberry$Market, sep = ""), rberry$Market)
rberry$Type <- ifelse(rberry$Type==" FRESH MARKET ", NA, rberry$Type)
rberry[is.na(rberry)] <- ""
rberry$Market <- ifelse(rberry$Type==" NOT HARVESTED ", paste(rberry$Type, rberry$Market, sep = ""), rberry$Market)
rberry$Type <- ifelse(rberry$Type==" NOT HARVESTED ", NA, rberry$Type)
rberry[is.na(rberry)] <- ""
rberry$Market <- ifelse(rberry$Type==" NOT SOLD ", paste(rberry$Type, rberry$Market, sep = ""), rberry$Market)
rberry$Type <- ifelse(rberry$Type==" NOT SOLD ", NA, rberry$Type)
rberry[is.na(rberry)] <- ""
rberry %<>% separate(Domain, c("D_left", "D_right"), sep = ", ")
rberry[is.na(rberry)] <- ""
rberry$Chemical_Type <- ifelse(rberry$D_left=="FERTILIZER", rberry$D_left, rberry$D_right)
rberry %<>% separate(`Domain Category`, c("DC_left", "DC_right"), sep = ":")
rberry[is.na(rberry)] <- ""
rberry %<>% rename(Chemicals = DC_right)
rberry %<>% select(-c(D_left, D_right, DC_left, Berry, Average))
kable(head(rberry)) %>% kable_styling(font_size = 12)
```

## Data Cleaning, Step 4

Remove NA's from the Value column.
```{r, echo=FALSE, warning=FALSE, message=FALSE}
## Remove NA's from Value
rberry$Value <- ifelse(rberry$Value=="(D)", NA, rberry$Value)
rberry[is.na(rberry)] <- ""
rberry$Value <- ifelse(rberry$Value=="(NA)", NA, rberry$Value)
rberry[is.na(rberry)] <- ""
kable(head(rberry)) %>% kable_styling(font_size = 12)
```

The dataset is now clean and ready for EDA. 

## Exploratory Data Analysis
- Goal: visualize the data and look for any patterns or trends
- For these analyses, only looking at values measured in lbs.
- First step: Histograms of Year and Number of lbs
```{r, echo=FALSE, warning=FALSE, message=FALSE}
rberry2 <- rberry %>% select(Year, State, Measure, Chemical_Type, Value)
rberry2 %<>% filter(Measure==" MEASURED IN LB")
rberry2$Value %<>% as.numeric()
hist_yrs <- hist(rberry2$Year, xlab = "Year", main = "Histogram of Years")
hist(rberry2$Value, xlab = "Number of LBs", main = "Histogram of Value (Measured in LB)")
```

## EDA
- Second step: scatterplot of Chemical Type vs. LBs
- points are jittered to avoid overlap
```{r, echo=FALSE, warning=FALSE, message=FALSE}
ggplot(data = rberry2, mapping = aes(Chemical_Type,(Value))) + geom_jitter(aes(color = State)) + ggtitle("Scatterplot of Chemical Type vs. LBs produced") + xlab("Chemical Type") + ylab("Value in LBs")
```

## EDA
- Third step: calculate mean number of lbs for various groups of observations
```{r, echo=FALSE, warning=FALSE, message=FALSE}
table1 <- rberry2 %>% group_by(State) %>% summarize(mean(Value, na.rm = TRUE))
table2 <- rberry2 %>% group_by(Year) %>% summarize(mean(Value, na.rm = TRUE))
table3 <- rberry2 %>% group_by(Chemical_Type) %>% summarise(mean(Value, na.rm = TRUE))
kable(table1) %>% kable_styling(font_size = 8)
kable(table2) %>% kable_styling(font_size = 8)
kable(table3) %>% kable_styling(font_size = 8)
```

## Conclusions
- Lots of NAs in the Value column
- 2016 and 2018 don't have many observations
- Distribution of lbs of raspberries is right-skewed
- "Other" categories don't have any observations
- No obvious patterns or trends in the data
- Future analyses may look at other measurement units besides lbs

## References
1. Hadley Wickham, Romain François, Lionel Henry and Kirill Müller
    (2020). dplyr: A Grammar of Data Manipulation. R package version
    1.0.2. https://CRAN.R-project.org/package=dplyr
2. Hao Zhu (2020). kableExtra: Construct Complex Table with 'kable' and
    Pipe Syntax. R package version 1.2.1.
    https://CRAN.R-project.org/package=kableExtra
3. H. Wickham. ggplot2: Elegant Graphics for Data Analysis.
    Springer-Verlag New York, 2016.
4. Stefan Milton Bache and Hadley Wickham (2014). magrittr: A
    Forward-Pipe Operator for R. R package version 1.5.
    https://CRAN.R-project.org/package=magrittr
5. Wickham et al., (2019). Welcome to the tidyverse. Journal of Open
    Source Software, 4(43), 1686, https://doi.org/10.21105/joss.01686
6. Yihui Xie (2020). knitr: A General-Purpose Package for Dynamic Report
    Generation in R. R package version 1.29.